### 2025-07-18 mythbusters.qmd
20250718 - summary text

In this chapter, we put a popular fantasy football belief to the test.
We evaluate whether the widely held belief is true that players perform better during a contract year.

@@

Then, we analyze the data.
Below is a mixed model that examines whether a player has a higher QBR per game when they are in a contract year compared to when they are not in a contract year.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

```{r}
# Placeholder for model predicting fantasy points
# Include player position as a covariate
```

@@

>>>>

In this chapter, we put a popular fantasy football belief to the test.
We evaluate the widely held belief that players perform better during a contract year.

@@

Then, we analyze the data.

#### Quarterback Rating {#sec-contractYearQB-QBR}

Below is a mixed model that examines whether a player has a higher QBR per game when they are in a contract year compared to when they are not in a contract year.
The first model includes just contract year as a predictor.
The second model includes additional covariates, including player age and experience.
In terms of Quarterback Rating (QBR), findings from the models indicate that Quarterbacks did not perform significantly better in their contract year.

@@

#### Points Added {#sec-contractYearQB-PointsAdded}

In terms of points added, Quarterbacks did not perform better in their contract year.

@@

#### Expected Points Added {#sec-contractYearQB-PointsAdded}

In terms of expected points added (EPA) from passing plays, when not controlling for player age and experience, Quarterbacks performed better in their contract year.
However, when controlling for player age and experience, Quarterbacks did not perform significantly better in their contract year.

@@

#### Fantasy Points {#sec-contractYearQB-FantasyPoints}

In terms of fantasy points, Quarterbacks performed significantly worse in their contract year, even controlling for player age and experience.

@@

#### Yards Per Carry {#sec-contractYearRB-YPC}

In terms of yards per carry (YPC), Running Backs did not perform significantly better in their contract year.

@@

#### Expected Points Added {#sec-contractYearRB-EPA}

In terms of expected points added (EPA) from rushing plays, Running Backs did not perform significantly better in their contract year.

@@

#### Fantasy Points {#sec-contractYearRB-FantasyPoints}

In terms of fantasy points, Running Backs performed significantly worse in their contract year, even controlling for player age and experience.

@@

#### Receiving Yards {#sec-contractYearWRTE-ReceivingYards}

In terms of receiving yards, Wide Receivers/Tight Ends performed significantly worse in their contract year, even controlling for player age and experience.

@@

#### Expected Points Added {#sec-contractYearWRTE-EPA}

In terms of expected points added (EPA) from receiving plays, Wide Receivers/Tight Ends performed significantly worse in their contract year, even controlling for player age and experience.

@@

#### Fantasy Points {#sec-contractYearWRTE-FantasyPoints}

In terms of fantasy points, Wide Receivers/Tight Ends performed significantly worse in their contract year, even controlling for player age and experience.

@@

#### Fantasy Points {#sec-contractYearQBRBWRTE-FantasyPoints}

In terms of fantasy points, Quarterbacks/Running Backs/Wide Receivers/Tight Ends performed significantly worse in their contract year, even controlling for player age and experience.

@@

@@

There is a widely held belief that NFL players perform better in the last year of the contract because they are motivated to gain another contract.
There is some evidence in the NBA and MLB that players tend to perform better in their contract year.
We evaluated this possibility among NFL players who were Quarterbacks, Running Backs, Wide Receivers, or Tight Ends.
We evaluated a wide range of performance indexes, including Quarterback Rating, yards per carry, points added, expected points added, receiving yards, and fantasy points.
None of the positions showed significantly better performance in their contract year for any of the performance indexes.
By contrast, if anything, players tended to perform more poorly during their contract year, as operationalized by fantasy points, receiving yards (WR/TE), and EPA from receiving plays (WR/TE), even when controlling for player and age experience.
In sum, we did not find evidence in support of the contract year hypothesis and consider this myth debunked.
However, we are open to this possibility being reexamined in new ways or with additional performance
